\documentclass{article}
\usepackage[utf8]{inputenc}
\begin{document}

\section*{OVERVIEW}
This meeting focused on fixed form and mean-field variational Bayes, with an emphasis on practical application, potential pitfalls (especially when estimating variance components and accounting for shrinkage), and strategies to clearly demonstrate these issues in example analyses for an introductory audience.

\section*{ACTIONITEMS}
\begin{itemize}
    \item Continue an extensive literature review on variational Bayes methods and approximation families.\\
    \textbf{Assignee:} David Ewing
    \item Implement and run two core example analyses: (1) demonstrate effects of incorrect independent blocking for strongly correlated parameters, and (2) hierarchical/variance-component examples contrasting low vs.\ high shrinkage (compare VB to MCMC).\\
    \textbf{Assignee:} David Ewing
    \item Create a focused example that explicitly shows the impact of placing strongly correlated parameters into separate independent blocks (under-dispersion).\\
    \textbf{Assignee:} David Ewing
    \item Prepare the written report and presentation explaining principles, examples, practical issues, and recommendations for using variational Bayes (targeted to an audience new to the method).\\
    \textbf{Assignee:} David Ewing
    \item Write up the results/draft and send the draft write-up to John for feedback.\\
    \textbf{Assignee:} David Ewing
    \item Schedule and confirm a Zoom meeting next week on a Monday or Wednesday (David’s availability noted).\\
    \textbf{Assignee:} David Ewing
    \item Plan and confirm an in-person meeting in Christchurch the week after next (John expects to be back then).\\
    \textbf{Assignee:} David Ewing
    \item Attend the follow-up Zoom meeting next week and the in-person meeting the week after next to discuss draft and examples.\\
    \textbf{Assignee:} John Holmes
\end{itemize}

\section*{SUMMARY}

\subsection*{Discussion on Fixed Form Variational Bayes}
\begin{itemize}
    \item John Holmes explains the concept of fixed form variational Bayes, emphasizing the specification of the variational distribution.
    \item David Ewing inquires about the mechanism for determining the best estimates of the parameters.
    \item John Holmes discusses the use of mean field variational Bayes and the identification of conditional conjugacy.
    \item The conversation touches on the minimization of the KL divergence and the use of the Stochastic Gradient Ascent algorithm for optimization.
\end{itemize}

\subsection*{Exploring Bayesian Logistic Regression and Random Effects Models}
\begin{itemize}
    \item John Holmes mentions a Bayesian logistic regression example and a random mental state model.
    \item David Ewing and John Holmes discuss the specifics of a random intercept model and the estimation of variance components.
    \item John Holmes highlights the sensitivity of the variance component distribution to the shrinkage of random effects.
    \item The conversation includes a discussion on the impact of non-flat priors on coefficient estimates and the introduction of shrinkage.
\end{itemize}

\subsection*{Challenges in Variational Bayes with Variance Components}
\begin{itemize}
    \item John Holmes explains the issues with variational Bayes in estimating variance components and the under-dispersion problem.
    \item David Ewing and John Holmes discuss the comparison of posterior variances from variational Bayes and MCMC methods.
    \item John Holmes emphasizes the importance of correctly specifying the degree of freedom in the true posterior distribution.
    \item The conversation includes a detailed explanation of the variational Bayes approach and its limitations in handling variance components.
\end{itemize}

\subsection*{Practical Considerations for Variational Bayes}
\begin{itemize}
    \item John Holmes advises on the practical considerations for fitting variational Bayes, including the importance of parameter block specification.
    \item David Ewing and John Holmes discuss the impact of strongly correlated parameters on inference accuracy.
    \item John Holmes suggests conducting examples to illustrate the advantages and disadvantages of variational Bayes.
    \item The conversation includes a plan for demonstrating variational Bayes with simple and variance component models.
\end{itemize}

\subsection*{Preparation for Report and Presentation}
\begin{itemize}
    \item John Holmes outlines the structure of the report, including the explanation of variational Bayes principles and examples.
    \item David Ewing expresses concerns about the depth of the report and the audience's understanding.
    \item John Holmes reassures David Ewing, emphasizing the need to explain the practical issues and advantages of variational Bayes.
    \item The conversation concludes with a plan for the next steps, including further reading and preparation for the presentation.
\end{itemize}

\end{document}